---
title: "Project 7: Difference-in-Differences and Synthetic Control"
output:
  pdf_document: default
  html_document: default
---


```{r}
# Install and load packages 
if (!require("pacman")) install.packages("pacman")

devtools::install_github("ebenmichael/augsynth")

pacman::p_load(# Tidyverse packages including dplyr and ggplot2 
               tidyverse,
               ggthemes,
               augsynth,
               gsynth)


library(readr)
library(tidyverse)

# set seed
set.seed(44)

# load data
medicaid_expansion <- read.csv('/Users/marisatsai/Library/CloudStorage/Box-Box/Spring 2024/CSSTP/projects/medicaid_expansion.csv')
summary(medicaid_expansion)

data<-medicaid_expansion
```

# Introduction

For this project, you will explore the question of whether the Affordable Care Act increased health insurance coverage (or conversely, decreased the number of people who are uninsured). The ACA was passed in March 2010, but several of its provisions were phased in over a few years. The ACA instituted the "individual mandate" which required that all Americans must carry health insurance, or else suffer a tax penalty. There are four mechanisms for how the ACA aims to reduce the uninsured population:

- Require companies with more than 50 employees to provide health insurance.
- Build state-run healthcare markets ("exchanges") for individuals to purchase health insurance.
- Provide subsidies to middle income individuals and families who do not qualify for employer based coverage.
- Expand Medicaid to require that states grant eligibility to all citizens and legal residents earning up to 138\% of the federal poverty line. The federal government would initially pay 100\% of the costs of this expansion, and over a period of 5 years the burden would shift so the federal government would pay 90\% and the states would pay 10\%.

In 2012, the Supreme Court heard the landmark case NFIB v. Sebelius, which principally challenged the constitutionality of the law under the theory that Congress could not institute an individual mandate. The Supreme Court ultimately upheld the individual mandate under Congress's taxation power, but struck down the requirement that states must expand Medicaid as impermissible subordination of the states to the federal government. Subsequently, several states refused to expand Medicaid when the program began on January 1, 2014. This refusal created the "Medicaid coverage gap" where there are indivudals who earn too much to qualify for Medicaid under the old standards, but too little to qualify for the ACA subsidies targeted at middle-income individuals.

States that refused to expand Medicaid principally cited the cost as the primary factor. Critics pointed out however, that the decision not to expand primarily broke down along partisan lines. In the years since the initial expansion, several states have opted into the program, either because of a change in the governing party, or because voters directly approved expansion via a ballot initiative.

You will explore the question of whether Medicaid expansion reduced the uninsured population in the U.S. in the 7 years since it went into effect. To address this question, you will use difference-in-differences estimation, and synthetic control.

# Data

The dataset you will work with has been assembled from a few different sources about Medicaid. The key variables are:

- **State**: Full name of state
- **Medicaid Expansion Adoption**: Date that the state adopted the Medicaid expansion, if it did so.
- **Year**: Year of observation.
- **Uninsured rate**: State uninsured rate in that year.

# Exploratory Data Analysis

Create plots and provide 1-2 sentence analyses to answer the following questions:

- Which states had the highest uninsured rates prior to 2014? The lowest?
- Which states were home to most uninsured Americans prior to 2014? How about in the last year in the data set? **Note**: 2010 state population is provided as a variable to answer this question. In an actual study you would likely use population estimates over time, but to simplify you can assume these numbers stay about the same.

```{r}
# highest and lowest uninsured rates


library(ggplot2)

# Filter data to include only years prior to 2014
filtered_data <- subset(data, year < 2014)

# Create the plot using ggplot2
ggplot(filtered_data, aes(x = year, y = uninsured_rate, group = State, color = State)) +
  geom_line() +
  labs(title = "Uninsured Rates by State Over Time (Before 2014)",
       x = "Year",
       y = "Uninsured Rate") +
  theme_minimal()



#getting list of states since it's hard to ID states by color 

# Filter data to include only year 2013
data_2013 <- subset(data, year == 2013)

# Calculate average uninsured rate for each state in 2013
avg_uninsured_rate_2013 <- aggregate(uninsured_rate ~ State, data_2013, mean)

# Sort states based on average uninsured rate in descending order
sorted_states <- avg_uninsured_rate_2013[order(-avg_uninsured_rate_2013$uninsured_rate), ]

# Print the sorted list of states
print(sorted_states)



```
Prior to 2014, DC and Massachusetts had the lowest uninsured rates; Texas, Nevada, and Florida had the highest.

```{r}
# most uninsured Americans

library(ggplot2)

# Filter data to include only years prior to 2014
filtered_data <- subset(data, year < 2014)

# Calculate number of uninsured people for each state
filtered_data$total_uninsured <- filtered_data$uninsured_rate * filtered_data$population

ggplot(filtered_data, aes(x = year, y = total_uninsured, group = State, color = State)) +
  geom_line() +
  labs(title = "Uninsured by State Over Time (Before 2014)",
       x = "Year",
       y = "Total uninsured") +
  theme_minimal()


#getting list of states since it's hard to ID states by color 

# Filter data to include only year 2013
data_2013 <- subset(data, year == 2013)

# Calculate number of uninsured people for each state in 2013
data_2013$total_uninsured <- data_2013$uninsured_rate * data_2013$population

# Aggregate total uninsured by state
total_uninsured_by_state <- aggregate(total_uninsured ~ State, data_2013, sum)

# Sort states based on total uninsured people in descending order
sorted_states_uninsured <- total_uninsured_by_state[order(-total_uninsured_by_state$total_uninsured), ]

# Print the sorted list of states
print(sorted_states_uninsured)

```

Prior to 2014, Vermont had the lowest number of uninsured; California had the highest.


# Difference-in-Differences Estimation

## Estimate Model

Do the following:

- Choose a state that adopted the Medicaid expansion on January 1, 2014 and a state that did not. **Hint**: Do not pick Massachusetts as it passed a universal healthcare law in 2006, and also avoid picking a state that adopted the Medicaid expansion between 2014 and 2015.
- Assess the parallel trends assumption for your choices using a plot. If you are not satisfied that the assumption has been met, pick another state and try again (but detail the states you tried).

```{r}
# Parallel Trends plot
library(dplyr)
# visualize intervention
# ----------
data %>%
  # processing 
  # ----------
  filter(State %in% c("California","Nebraska")) %>%  # use "%in% to filter values in a vector
  filter(year >= 2007 & year< 2014) %>%  
  #filter(between(year_qtr, 2012.5, 2012.75)) %>%   # same filtering but using between() instead which I find simpler
  
  # plot
  # ----------
  ggplot() + 
  # add in point layer
  geom_point(aes(x = year, 
                 y = uninsured_rate, 
                 color = State)) + # color by state
  # add in line
  geom_line(aes(x = year, 
                y = uninsured_rate, 
                color = State)) +
  
  # themes
 # theme_fivethirtyeight() +
  #theme(axis.title = element_text()) +
  
  # labels - PREFER TO USE labs() SO THAT IT IS ALL IN ONE ARGUMENT
  ggtitle('CA and NE uninsured rate before 2014 Medicaid expansion') +
  xlab('Year') +
  ylab('Uninsured rate')


```
I tried California and Utah, but they were extremely different - Utah experienced a big drop in 2010. I then tried NE, which was more similar to each other, in that both increased in 2010/2011 and decreased afterwards.

- Estimate a difference-in-differences estimate of the effect of the Medicaid expansion on the uninsured share of the population. You may follow the lab example where we estimate the differences in one pre-treatment and one post-treatment period, or take an average of the pre-treatment and post-treatment outcomes

```{r}
# Difference-in-Differences estimation

#
# DiD for: CA-NE
# ----------------------------------------------------------------
# create a dataset for CA and NE
kc <- 
  data %>%
  filter(State %in% c("California","Nebraska")) %>%
  filter(year >= 2013 & year<= 2014) 

# pre-treatment difference
# ----------

library(tidyr)
pre_diff <- 
  kc %>%
  # filter out only the quarter we want
  filter(year == 2013) %>%
  # subset to select only vars we want
  select(State, 
         uninsured_rate) %>%
  # make the data wide
  pivot_wider(names_from = State, 
              values_from = uninsured_rate) %>%
  #subtract to make calculation
  summarise(diff = Nebraska - California)
  
# post-treatment difference
# ----------
post_diff <- 
  kc %>%
  # filter out only the quarter we want
  filter(year == 2014) %>%
  # subset to select only vars we want
  select(State, 
         uninsured_rate) %>%
  # make the data wide
  pivot_wider(names_from = State, 
              values_from = uninsured_rate) %>%
  # subtract to make calculation
  summarise(Nebraska - California)

# diff-in-diffs
# ----------
diff_in_diffs <- post_diff - pre_diff
diff_in_diffs

```


## Discussion Questions

- Card/Krueger's original piece utilized the fact that towns on either side of the Delaware river are likely to be quite similar to one another in terms of demographics, economics, etc. Why is that intuition harder to replicate with this data?
- **Answer**: It is harder to replicate because our data represents state, which has larger variation in geography, demographic make-up, and surrounding policies than towns across a river.

- What are the strengths and weaknesses of using the parallel trends assumption in difference-in-differences estimates?
- **Answer**: Having parallel trends makes it more conceiveable that a deviation from the trend at the timepoint of an intervention might be due to the intervention; it suggests that factors (outside of the intervention) affecting trends between the two groups might be consistent. A weakness is that people may overinterpret differences since there could still be factors that cause deviations from trends that are not due to an interbentions.  


# Synthetic Control

Estimate Synthetic Control

Although several states did not expand Medicaid on January 1, 2014, many did later on. In some cases, a Democratic governor was elected and pushed for a state budget that included the Medicaid expansion, whereas in others voters approved expansion via a ballot initiative. The 2018 election was a watershed moment where several Republican-leaning states elected Democratic governors and approved Medicaid expansion. In cases with a ballot initiative, the state legislature and governor still must implement the results via legislation. For instance, Idaho voters approved a Medicaid expansion in the 2018 election, but it was not implemented in the state budget until late 2019, with enrollment beginning in 2020.

Do the following:

- Choose a state that adopted the Medicaid expansion after January 1, 2014. Construct a non-augmented synthetic control and plot the results (both pre-treatment fit and post-treatment differences). Also report the average ATT and L2 imbalance.

```{r}


# create a treatment indicator 
# ----------
data <- 
  data %>%
  # select subset of variables 
  select(year, State, Date_Adopted, population, uninsured_rate) %>%
  # create new treatment flag just to see
  mutate(treatment = case_when(State == "Nebraska" & year >= 2013 ~ 1, 
                               TRUE ~ 0))
# view head
head(data)

# install libraries - install "augsynth" here since it is now on CRAN
pacman::p_load(# Tidyverse packages including dplyr and ggplot2 
               tidyverse,
               ggthemes,
               augsynth)  
#install.packages('BiocManager')
#library(augsynth)
#install.packages("augsynth")
#install.packages("devtools")
#devtools::install_github("ebenmichael/augsynth")


# non-augmented synthetic control

  syn <-                              # save object 
  augsynth(uninsured_rate ~ treatment, # treatment
                         State,     # unit
                         year,  # time
                         data,    # data
           progfunc = "None",       # plain syn control
           scm = T)                 # synthetic control 

# summary 
summary(syn)
```
```{r}
# plot
plot(syn)
```

```

- Re-run the same analysis but this time use an augmentation (default choices are Ridge, Matrix Completion, and GSynth). Create the same plot and report the average ATT and L2 imbalance.

```{r}
# augmented synthetic control
ridge_syn <-
augsynth(uninsured_rate ~ treatment,
State,
year,
data,
progfunc = "ridge", 
scm = T)

# summary 
summary(ridge_syn)

```

- Plot barplots to visualize the weights of the donors.

```{r}
# barplots of weights
data.frame(ridge_syn$weights) %>%
  tibble::rownames_to_column('State') %>%
  ggplot() +
  geom_bar(aes(x = State, y = ridge_syn.weights),
           stat = 'identity') +
  coord_flip() + # coord flip
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('Synthetic Control Weights') +
  xlab('State') +
  ylab('Weight') 
```

**HINT**: Is there any preprocessing you need to do before you allow the program to automatically find weights for donor states?

## Discussion Questions


- What are the advantages and disadvantages of synthetic control compared to difference-in-differences estimators?
- **Answer**: 

- One of the benefits of synthetic control is that the weights are bounded between [0,1] and the weights must sum to 1. Augmentation might relax this assumption by allowing for negative weights. Does this create an interpretation problem, and how should we balance this consideration against the improvements augmentation offers in terms of imbalance in the pre-treatment period?
- **Answer**: I don't think it creates an interpretation issue-- in weighting, it seems natural to have both up- and down-weighted influences. I don't see a problem if it improves balance issues.

# Staggered Adoption Synthetic Control

## Estimate Multisynth

Do the following:

- Estimate a multisynth model that treats each state individually. Choose a fraction of states that you can fit on a plot and examine their treatment effects.

```{r}
# multisynth model states
summary(data)

# create dataset
# ---------
data_clean <- 
  data %>%
#create "treatment" - year Medicaid expansion was adopted
    mutate(Date_Adopted = ifelse(is.na(Date_Adopted), 
                                   Inf, Date_Adopted),
           DA = 1 * (year >= Date_Adopted))
```

```{r}

# setting nu to 0.5
# ---------
ppool_syn <- multisynth(uninsured_rate ~ DA, 
                        State,                       # unit
                        year,                        # time
                        nu = 0,                    # varying degree of pooling 
                        data_clean, # data
                        n_leads = 3)                # post-treatment periods to estimate 


# view results 
print(ppool_syn$nu)

ppool_syn
```

```{r}
# save ATT and balance stats
# ---------
ppool_syn_summ <- summary(ppool_syn)
```

```{r}
# plot actual estimates not values of synthetic controls
# ---------
ppool_syn_summ$att %>%
  ggplot(aes(x = Time, y = Estimate, color = Level)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        legend.position = "bottom") +
  ggtitle('Synthetic Controls for Medicaid expansion') +
  xlab('Time') +
  ylab('uninsured rate')

```


```{r}

# plot actual estimates not values of synthetic controls
# ---------

ppool_syn_summ$att %>%
  ggplot(aes(x = Time, y = Estimate, color = Level)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        legend.position = 'None') +
  ggtitle('Synthetic Controls for Medicaid expansion') +
  xlab('Time') +
  ylab('Uninsured rate') +
  facet_wrap(~Level)   # facet-wrap by level (state in this case) for clearer presentation
```



- Estimate a multisynth model using time cohorts. For the purpose of this exercise, you can simplify the treatment time so that states that adopted Medicaid expansion within the same year (i.e. all states that adopted epxansion in 2016) count for the same cohort. Plot the treatment effects for these time cohorts.

```{r}
# multisynth model time cohorts

summary(data)
# break observations into time cohorts 
# ----------------------------------------------------------------
ppool_syn_time <- multisynth(uninsured_rate ~ DA,
                             State, 
                             year,
                             data_clean, 
                             n_leads = 6, 
                             time_cohort = TRUE)           # time cohort set to TRUE

# save summary
ppool_syn_time_summ <- summary(ppool_syn_time)

# view 
ppool_syn_time_summ
```

```{r}

# plot effect for each time period (local treatment effects)
# ---------
ppool_syn_time_summ$att %>%
  ggplot(aes(x = Time, y = Estimate, color = Level)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        legend.position = 'None') +
  ggtitle('Synthetic Controls for Medicaid expansion') +
  xlab('Time') +
  ylab('Uninsured rate') +
  facet_wrap(~Level)
```


## Discussion Questions

- One feature of Medicaid is that it is jointly administered by the federal government and the states, and states have some flexibility in how they implement Medicaid. For example, during the Trump administration, several states applied for waivers where they could add work requirements to the eligibility standards (i.e. an individual needed to work for 80 hours/month to qualify for Medicaid). Given these differences, do you see evidence for the idea that different states had different treatment effect sizes?
- **Answer**: Yes, the trend does not look entirely consistent across states.

- Do you see evidence for the idea that early adopters of Medicaid expansion enjoyed a larger decrease in the uninsured population?
- **Answer**: No, that wasn't clear in the plots.

# General Discussion Questions

- Why are DiD and  synthetic control estimates well suited to studies of aggregated units like cities, states, countries, etc?
- **Answer**: DiD is often used to analyze policy-level changes which occur in aggregated unit. Policy-level changes are often focused on changing some aggregate level outcome, over time so having an method that can incorporate the unit-level and longituinal nature is helpful. Aggregated units may not have a siple 

- What role does selection into treatment play in DiD/synthetic control versus regression discontinuity? When would we want to use either method?
- **Answer**: In DiD, selection into treatment is on an aggregate scale; in regression discontinuity it is individual, based on a cutoff of some metric.